---
title: "Analyzing Forest Fires"
output: html_notebook
---

```{r}
require(tidyverse)
require(ggplot2)

forestfires <- read.csv("forestfires.csv")
head(forestfires, 50)
```
Loading tidyverse and ggplot2 packages and importing the forest fire csv file into a dataframe. We use head to look at the first several rows in order to get a sense of the structure and formatting of the data.
```{r}
forestfires %>% pull(month) %>% unique()
forestfires %>% pull(day) %>% unique()
```
Taking a look at the month and day variables we can see how they are abbreviated as well as how they are ordered.
```{r}
forestfires <- forestfires %>%
    mutate(
      month = factor(month, levels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")),
      day = factor(day, levels = c("mon", "tue", "wed", "thu", "fri", "sat", "sun"))
    )
```
We can change the order of the month and day variables as well as changing them into categorical variables.
```{r}
numff.month <- forestfires %>%
  group_by(month) %>%
  summarize(
   n = n()
  )
numff.day <- forestfires %>%
  group_by(day) %>%
  summarize(
   n = n()
  )
numff.month %>%
      ggplot(aes(x = month, y = n)) +
      geom_col() +
      labs(
        title = "Forest Fires by Month",
        y = "Number of Forest Fires")
numff.day %>%
      ggplot(aes(x = day, y = n)) +
      geom_col() +
      labs(
        title = "Forest Fires by Day of the Week",
        y = "Number of Forest Fires")
```
We group the data by both month and day, then summarize the dataframes by counting the number of observations for each month/day and load those summaries into a new data frame, respectively. Thanks to the categoricalization of the variables 'month' and 'day' we are able to have the designated order in our graphs. 

Quick takeaways; Peak forest fire season is within the months of the August and September. Additionally, forest fires tend to occur during the weekend i.e. Friday-Sunday. 
```{r}
forestfires_long <- forestfires %>%
  pivot_longer(
    cols = c(FFMC, DMC, DC, ISI, temp, RH, wind, rain),
    names_to = "variables",
    values_to = "values"
  )
forestfires_long %>%
  ggplot(aes(x=month, y=values)) +
  geom_line() +
  facet_wrap(vars(variables),
  scales = "free_y")

forestfires_long %>%
  ggplot(aes(x=day, y=values)) +
  geom_line() +
  facet_wrap(vars(variables),
  scales = "free_y")
```
Values for the following indices have positive correlation with the number of forest fires by month; DC, DMC, and Temperature. This shows that Temperature, Rain, and Relative humidity are significantly contributing factors as they factor into the calculation for DC, and DMC. 
```{r}
forestfires_long %>%
  ggplot(aes(x=values, y=area)) +
  geom_point() +
  facet_wrap(vars(variables) ,
  scales = "free_x")

cor_ff <- cor(forestfires[13], forestfires[5:12], method = "spearman")
head(cor_ff)
```
Higher values of FFMC, DC, and Temperature show more correlation with higher values of forest fire area (intensity). Running a correlative comparison shows temperature as having the highest p-value.
```{r}
forestfires_long_filter <- forestfires_long %>%
  filter(area > 0)
forestfires_long_filter %>%
  ggplot(aes(x=values, y=log2(area))) +
  geom_point() +
  facet_wrap(vars(variables) ,
  scales = "free_x")
  
```

