---
title: "Dealing With Missing Data"
author: "DDunn"
date: '2023-05-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this project we will be examining missing data in a NYC motor vehicle collisions dataset for 2018 sourced from the NYC OpenData website. We load the csv file into a dataframe and then create a matrix which counts the na values in each of the variables of the dataset. 

```{r}
library(tidyverse)
mvc <- read.csv("nypd_mvc_2018.csv")
na_logical <- is.na(mvc)
```

Creating a dataframe which contains logical values that indicate where the missing values are. Then creating a new column which accounts for the total number of injured; to do so we will rowSum to compute the number of NA and use mutate to add the new column.

```{r}
mvc_na <- data.frame(na_logical)
mvc_na_injured <- mvc_na %>%
  mutate(total_na_injured = rowSums(.[9:11]))
print(mvc_na_injured)
```

The dataframe "killed" contains the four columns that relate to people killed in collisions. Filtering for the killed rows where each value in the manual_sum column either is not equal to the values in the total_killed column or the total_killed column is NA value.

```{r}
killed  <-  mvc %>% select(ends_with("_killed"))
killed_non_eq <- killed %>%
    mutate(manual_sum = rowSums(.[1:3])) %>%
    filter(manual_sum != total_killed | is.na(total_killed))
```

Cleaning the killed_non_eq and injured_non_eq columns by replacing any NA values with their corresponding values from the manual_sum column. We then use the same combination to replace any values that aren't equal to their corresponding values in manual_sum with NaN values.

```{r}
# fix the killed values
killed_non_eq <- killed_non_eq %>%
    mutate(total_killed = if_else(is.na(total_killed), manual_sum, total_killed ))

killed_non_eq <- killed_non_eq %>%
    mutate(total_killed = if_else(total_killed != manual_sum, NaN, total_killed ))

# Create an injured_non_eq dataframe and manually sum values
injured  <-  mvc %>% select(ends_with("_injured"))

injured_non_eq <- injured %>% 
    mutate(manual_sum = rowSums(.[1:3])) %>%
    filter(manual_sum != total_injured | is.na(total_injured))

injured_non_eq <- injured_non_eq %>%
    mutate(total_injured = if_else(is.na(total_injured), manual_sum, total_injured)) %>%
    mutate(total_injured = if_else(total_injured != manual_sum, NaN, total_injured))
```

Using the pivot_longer function we transform the mvc_na dataframe into a new dataframe containing x and value variables. Using the everything() function to indicate that all columns are involved in the reshaping process. We then create a new column y containing the row numbers.

```{r}
mvc  <-  read_csv("nypd_mvc_2018.csv")
mvc_na <- map_df(mvc, function(x) as.numeric(is.na(x)))

mvc_na_heat <- mvc_na %>%
    pivot_longer(cols = everything(),
                 names_to = "x",
                 values_to = "value") %>%
    group_by(x) %>%
    mutate(y = row_number())
```

We have a plotting function which will create a heatmap. We use this function on a subset of the mvc_na dataframe where variables for vehicles are chosen.

```{r}
plot_na_matrix <- function(df) {
    # Preparing the dataframe for heatmaps 
    df_heat <- df %>%
        pivot_longer(cols = everything(),
               names_to = "x") %>%
        group_by(x) %>%
        mutate(y = row_number())
    
    # Ensuring the order of columns is kept as it is
    df_heat <- df_heat %>%
        ungroup() %>%
        mutate(x = factor(x,levels = colnames(df)))
    
    # Plotting data
    g <- ggplot(data = df_heat, aes(x=x, y=y, fill=value)) + 
        geom_tile() + 
        theme(legend.position = "none",
              axis.title.y=element_blank(),
              axis.text.y =element_blank(),
              axis.ticks.y=element_blank(),
              axis.title.x=element_blank(),
              axis.text.x = element_text(angle = 90, hjust = 1))
    
    # Returning the plot
    g
}

mvc_na_vehicle <- mvc_na %>%
    select(contains("vehicle"))
plot_na_matrix(mvc_na_vehicle)
```

Setting up and defining the plotting function which will create a heatmap when given a correlation matrix. We use the contains function to select columns from the mvc_na dataframe that contain the substring 'vehicle'. Then creating the correlation matrix and rounding the result to 2 digits after the decimal point we pass this into the heatmap plotting function.

```{r}
#plot na correlation helper
plot_na_correlation <- function(df) {
    # Taking the lower triangle of the correlation matrix
    missing_corr_up <- df
    missing_corr_up[lower.tri(missing_corr_up)] <- NA
    missing_corr_up <- data.frame(missing_corr_up)
    
    # Preparing the dataframe for heatmaps 
    col_names <- colnames(missing_corr_up)
    
    missing_corr_up_heat <- missing_corr_up %>%
        pivot_longer(cols = everything(),
               names_to = "x") %>%
        group_by(x) %>%
        mutate(y = col_names[row_number()])  %>%
        na.omit
    
    # Ordering triangle
    ordered_cols_asc <- col_names[order(colSums(is.na(missing_corr_up)))]
    ordered_cols_desc <- col_names[order(-colSums(is.na(missing_corr_up)))]
    
    missing_corr_up_heat <- missing_corr_up_heat %>%
        ungroup() %>%
        mutate(x = factor(x,levels = ordered_cols_asc)) %>%
        mutate(y = factor(y,levels = ordered_cols_desc))
    
    # Plotting heatmaps
    g <- ggplot(data = missing_corr_up_heat, aes(x=x, y=y, fill=value)) + 
        geom_tile() + 
        geom_text(aes(label=value)) +
        theme_minimal() +
        scale_fill_gradientn(colours = c("white", "yellow", "red"), values = c(-1,0,1)) +
        theme(legend.position = "none",
              axis.title.y=element_blank(),
              axis.title.x=element_blank(),
              axis.text.x = element_text(angle = 90, hjust = 1))
    
    # Returning the plot
    g
}
mvc_na_vehicle <- mvc_na %>%
    select(contains("vehicle"))
missing_vehicle_corr <- round(cor(mvc_na_vehicle),2)
plot_na_correlation(missing_vehicle_corr)
    
```

A function v_fun() to count the number of rows where the column vehicle_x is NA and the column cause_vehicle_x is not NA (x is the number of vehicle and cause vehicle). A variable v_na containing the number of vehicle missing using map_int() and v_fun() functions. Variable c_na contains the number of cause missing. Function c_fun() counts the number of rows where the column vehicle_x is not NA and the column cause_vehicle_x is NA (x being the vehicle number).

```{r}
col_labels  <-  c('v_number', 'vehicle_missing', 'cause_missing')

v_fun <- function(x){ 
    v_col <- paste('vehicle', x,  sep = "_" )
    c_col <- paste('cause_vehicle', x,  sep = "_" )
    sum(is.na(mvc[v_col]) & !is.na(mvc[c_col]))
}
v_na <- map_int(1:5, v_fun )

c_fun <- function(x){
    v_col <- paste('vehicle', x, sep = "_")
    c_col <- paste('cause_vehicle', x, sep = "_")
    sum(is.na(mvc[c_col]) & !is.na(mvc[v_col]))
}
c_na <- map_int(1:5, c_fun)
#v <- c(1, 2, 3, 4 ,5)
vc_na_df <- tibble(1:5, v_na, c_na)
colnames(vc_na_df) <- col_labels
```

The dataframe vehicles only contains the columns from mvc starting with "vehicle". Using the pivot_longer() function we stack the values form the vehicles dataframe in to a key-value dataframe. Then using the table() function we count the unique values from the stacked list. 

```{r}
vehicles  <-  mvc %>% select(starts_with("vehicle_"))
head(vehicles)
vehicles_longer <- vehicles %>% pivot_longer(cols = everything(),
                                names_to = "vehicle_num",
                                values_to = "vehicle_type")
top10_vehicles <- head(sort(table(vehicles_longer$vehicle_type), decreasing = T), 10)
print(top10_vehicles)
```

The helper function summarize_missing contains the logic to count missing values across the pairs of columns. The for loop creates al ogical vector for values where the vehicle column is NA and the cause column is non-NA; it creates another logical vector where the reverse is true. The loop also used the first logical vector to fill matching values from the vehicle AND the cause columns with the string 'Unspecified'. The other helper functions v_fun and c_fun take an integer input and calculate the number of missing values in the vehicle and cause columns where the corresponding cause and vehicle column are not missing, respectively.

```{r}
v_fun <- function(x){ 
    v_col <- paste('vehicle', x,  sep = "_" )
    c_col <- paste('cause_vehicle', x,  sep = "_" )
    sum(is.na(mvc[v_col]) & !is.na(mvc[c_col]))
}

c_fun <- function(x){ 
    v_col <- paste('vehicle', x,  sep = "_" )
    c_col <- paste('cause_vehicle', x,  sep = "_" )
    sum(!is.na(mvc[v_col]) & is.na(mvc[c_col]))
}

summarize_missing <- function(){
    library(purrr)
    
    col_labels  <-  c('v_number', 'vehicle_missing', 'cause_missing')
    
    v_na <- map_int(1:5, v_fun )
    c_na <- map_int(1:5, c_fun )
    
    vc_na_df  <-  tibble(1:5, v_na, c_na)
    colnames(vc_na_df) <- col_labels
    vc_na_df
}

summary_before  <-  summarize_missing() 

 for (x in 1:5 ){
    v_col <- paste('vehicle', x,  sep = "_" )
    c_col <- paste('cause_vehicle', x,  sep = "_" )
    v_na_logic <- is.na(mvc[v_col]) & !is.na(mvc[c_col])
    c_na_logic <- !is.na(mvc[v_col]) & is.na(mvc[c_col])
    mvc <- mvc %>%
        mutate_at(c(v_col), function(x) 
        if_else(v_na_logic, "Unspecified",x)) %>%
        mutate_at(c(c_col), function(x) 
        if_else(c_na_logic, "Unspecified",x))
}

summary_after <- summarize_missing()
```

Looping over the column names in location_cols, for each iteration of the loop we're replacing values in the column in the mvc. The logical vector represents whether the values in the column in the mvc dataframe has an NA value or not. WHere the logical vector is true the value is replaced with the equivalent value from sup_data. The number of NA values across location_cols in mvc is calculated and assigned to na_after.

```{r}
sup_data  <-  read_csv('supplemental_data.csv')

location_cols  <-  c('location', 'on_street', 'off_street', 'borough')
na_before  <-  colSums(is.na(mvc[location_cols]))

for (col in location_cols){
    na_logic <- is.na(mvc[col])
    mvc[na_logic, col] <- sup_data[na_logic, col]
    }
na_after <- colSums(is.na(mvc[location_cols]))
```

Ideas for continued work with the motor vehicle collision data set:

1. Drop the rows that had suspect values for injured and killed totals.

2. Clean the values in the vehicle_1 through vehicle_5 columns by analyzing the different values and merging duplicates and near-duplicates.

3. Analyze whether collisions are more likely in certain locations, at certain times, or for certain vehicle types.


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
